
---
title: "Exposure Calibration - Saguenay Fjord"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# 1. Load Libraries
```{r libraries}
library(raster)
library(sf)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(corrplot)

```

# 2. Load Benthic Data with Shannon Index
```{r load_benthic}
# Load the benthic data we created earlier
benthic_data <- read.csv("benthic_data_complete.csv")

# Convert to spatial object
benthic_sf <- st_as_sf(benthic_data,
                       coords = c("Longitude", "Latitude"),
                       crs = 4326)

# Transform to UTM 19N (same as exposure rasters)
benthic_sf <- st_transform(benthic_sf, crs = 32619)

# Summary
cat("Number of stations:", nrow(benthic_sf), "\n")
cat("Shannon Index range:", range(benthic_sf$Shannon_Index), "\n")

# Preview
head(benthic_sf)
```

# 3. Define Stressors and Coefficients
```{r define_parameters}
# List of stressors (based on your files)
stressors <- c(
  "agriculture",
  "coastal_development",
  "ice_fishing",
  "wwtp",
  "forestry"
)

# Coefficients to test
coefficients <- c(0.00001, 0.00005, 0.000025)

cat("Stressors:", length(stressors), "\n")
cat("Coefficients to test:", coefficients, "\n")
```

# 4. Load Exposure Rasters Function

```{r load exposure}
# ================================================================
# CORRECTED FUNCTION - Fixed coastal_development and rounding
# ================================================================

load_exposure_raster <- function(stressor, coefficient) {
  
  # Convert to string with ENOUGH decimal places
  # Use 6 decimals to preserve 0.000025
  coef_str <- sprintf("%.6f", coefficient)  # Changed from %.5f to %.6f
  
  # Remove trailing zeros but keep at least 5 decimals for 0.00001
  coef_str <- gsub("0+$", "", coef_str)  # Remove trailing zeros
  if (!grepl("\\.", coef_str)) {
    coef_str <- paste0(coef_str, ".0")
  }
  
  # For ice_fishing and coastal_development (no "0." prefix)
  coef_str_short <- gsub("^0+\\.", "", sprintf("%.6f", coefficient))
  coef_str_short <- gsub("0+$", "", coef_str_short)
  
  cat("    Looking for coefficient format:", coef_str, "(short:", coef_str_short, ")\n")
  
  # Try different file naming patterns based on stressor
  file_path <- NULL
  
  if (stressor == "agriculture") {
    # Pattern: agriculture_exposure.0.00001.tif
    file_path <- sprintf("agriculture_exposure.%s.tif", coef_str)
    
  } else if (stressor == "coastal_development") {
    # Pattern: coastal_development_exposure.00001.tif (NO "0." prefix, like ice_fishing)
    file_path <- sprintf("coastal_development_exposure.%s.tif", coef_str_short)
    
  } else if (stressor == "ice_fishing") {
    # Pattern: ice_fishing_exposure.00001.tif (no "0." prefix)
    file_path <- sprintf("ice_fishing_exposure.%s.tif", coef_str_short)
    
  } else if (stressor == "wwtp") {
    # Pattern: wwtp_exposure_0.00001.tif (underscore before coefficient)
    file_path <- sprintf("wwtp_exposure_%s.tif", coef_str)
    
  } else if (stressor == "forestry") {
    # Pattern: forestry_exposure.0.00001.RData
    file_path <- sprintf("forestry_exposure.%s.RData", coef_str)
  }
  
  # Print what we're looking for
  cat("    Trying to load:", file_path, "\n")
  
  # Check if file exists
  if (file.exists(file_path)) {
    
    if (stressor == "forestry") {
      # Load .RData file
      env <- new.env()
      load(file_path, envir = env)
      
      # Try to find the raster object
      obj_names <- ls(envir = env)
      
      # Look for raster objects
      for (obj_name in obj_names) {
        obj <- get(obj_name, envir = env)
        if (inherits(obj, c("RasterLayer", "RasterStack", "RasterBrick"))) {
          cat("    ✓ Found raster in .RData:", obj_name, "\n")
          return(obj)
        }
      }
      
      warning(paste("No raster found in", file_path))
      return(NULL)
      
    } else {
      # Load .tif file
      return(raster(file_path))
    }
    
  } else {
    cat("    ✗ File not found:", file_path, "\n")
    
    # Try listing similar files to help debug
    pattern <- paste0("^", stressor)
    similar_files <- list.files(pattern = pattern)
    
    if (length(similar_files) > 0) {
      cat("    Similar files found:", paste(head(similar_files, 5), collapse = ", "), "\n")
    }
    
    return(NULL)
  }
}
```


# 5. Calculate Cumulative Exposure for Each Coefficient

```{r calculate_cumulative }
# Initialize results dataframe
results <- data.frame(
  Coefficient = numeric(),
  Correlation_Spearman = numeric(),
  Correlation_Pearson = numeric(),
  R_squared = numeric(),
  P_value = numeric()
)

# Store cumulative rasters for later visualization
cumulative_rasters <- list()

# Loop through each coefficient
for (coef in coefficients) {
  
  cat("\n========================================\n")
  cat("Processing coefficient:", coef, "\n")
  cat("========================================\n")
  
  # Initialize cumulative exposure raster
  cumulative_exposure <- NULL
  loaded_count <- 0
  
  # Load and sum all stressor exposures
  for (stressor in stressors) {
    
    exposure_rast <- load_exposure_raster(stressor, coef)
    
    if (!is.null(exposure_rast)) {
      
      if (is.null(cumulative_exposure)) {
        # First raster - initialize
        cumulative_exposure <- exposure_rast
      } else {
        # Add to cumulative
        cumulative_exposure <- cumulative_exposure + exposure_rast
      }
      
      loaded_count <- loaded_count + 1
      cat("  ✓ Loaded", stressor, "\n")
      
    } else {
      cat("  ✗ Failed to load", stressor, "\n")
    }
  }
  
  cat("  Total stressors loaded:", loaded_count, "out of", length(stressors), "\n")
  
  if (!is.null(cumulative_exposure)) {
    
    # Normalize to 0-1 range
    max_val <- cellStats(cumulative_exposure, "max", na.rm = TRUE)
    if (!is.na(max_val) && max_val > 0) {
      cumulative_exposure <- cumulative_exposure / max_val
    }
    
    # Store for later
    cumulative_rasters[[as.character(coef)]] <- cumulative_exposure
    
    # Extract at benthic stations - USE raster::extract() EXPLICITLY
    extracted_values <- raster::extract(cumulative_exposure, st_coordinates(benthic_sf))
    
    # Add to benthic data
    benthic_sf[[paste0("Exposure_", coef)]] <- extracted_values
    
    # Calculate correlations
    valid_idx <- complete.cases(extracted_values, benthic_sf$Shannon_Index)
    
    if (sum(valid_idx) > 10) {
      
      spearman_cor <- cor(extracted_values[valid_idx],
                          benthic_sf$Shannon_Index[valid_idx],
                          method = "spearman")
      
      pearson_cor <- cor(extracted_values[valid_idx],
                         benthic_sf$Shannon_Index[valid_idx],
                         method = "pearson")
      
      # Linear model
      model <- lm(Shannon_Index ~ extracted_values, 
                  data = data.frame(Shannon_Index = benthic_sf$Shannon_Index[valid_idx],
                                   extracted_values = extracted_values[valid_idx]))
      
      r_squared <- summary(model)$r.squared
      p_value <- summary(model)$coefficients[2, 4]  # P-value for slope
      
      # Store results
      results <- rbind(results, data.frame(
        Coefficient = coef,
        Correlation_Spearman = spearman_cor,
        Correlation_Pearson = pearson_cor,
        R_squared = r_squared,
        P_value = p_value
      ))
      
      cat("  Spearman correlation:", round(spearman_cor, 4), "\n")
      cat("  Pearson correlation:", round(pearson_cor, 4), "\n")
      cat("  R²:", round(r_squared, 4), "\n")
      cat("  P-value:", format.pval(p_value), "\n")
      
    } else {
      cat("  ✗ Not enough valid data points for correlation\n")
    }
    
  } else {
    cat("  ✗ Failed to create cumulative exposure\n")
  }
}
```

# 6. Results Summary
```{r results_table}
# Sort by absolute Spearman correlation
results <- results %>%
  mutate(Abs_Correlation = abs(Correlation_Spearman)) %>%
  arrange(desc(Abs_Correlation))

# Display results table
kable(results, digits = 4, caption = "Calibration Results - All Coefficients")

# Best coefficient
best_coef <- results$Coefficient[1]
best_cor <- results$Correlation_Spearman[1]

cat("\n========================================\n")
cat("BEST COEFFICIENT:", best_coef, "\n")
cat("SPEARMAN CORRELATION:", round(best_cor, 4), "\n")
cat("========================================\n")
```

# 7. Visualization: Correlation by Coefficient
```{r plot_correlation_curve}
ggplot(results, aes(x = Coefficient, y = Correlation_Spearman)) +
  geom_line(size = 1.5, color = "steelblue") +
  geom_point(size = 4, color = "darkblue") +
  geom_point(data = results %>% filter(Coefficient == best_coef),
             size = 6, color = "red", shape = 21, fill = "yellow") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Calibration Curve: Correlation vs Coefficient",
    subtitle = paste("Best coefficient:", best_coef, "| Correlation:", round(best_cor, 4)),
    x = "Coefficient Value",
    y = "Spearman Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank()
  )
```

# 8. Scatter Plots: Shannon Index vs Exposure
```{r scatter_plots, fig.height=10}
# Prepare data for plotting
plot_data <- benthic_sf %>%
  st_drop_geometry() %>%
  select(Station, Shannon_Index, starts_with("Exposure_")) %>%
  pivot_longer(cols = starts_with("Exposure_"),
               names_to = "Coefficient",
               values_to = "Exposure") %>%
  mutate(Coefficient = gsub("Exposure_", "", Coefficient))

# Create scatter plots for each coefficient
ggplot(plot_data, aes(x = Exposure, y = Shannon_Index)) +
  geom_point(size = 3, alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", fill = "pink", alpha = 0.3) +
  facet_wrap(~Coefficient, ncol = 1, scales = "free_x",
             labeller = labeller(Coefficient = function(x) paste("Coefficient:", x))) +
  labs(
    title = "Shannon Index vs Cumulative Exposure",
    subtitle = "Each panel shows a different coefficient value",
    x = "Cumulative Exposure (normalized 0-1)",
    y = "Shannon Diversity Index"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "lightblue"),
    strip.text = element_text(face = "bold")
  )
```

# 9. Best Model Diagnostics
```{r best_model_diagnostics}
# Extract data for best coefficient
best_exposure <- benthic_sf[[paste0("Exposure_", best_coef)]]
valid_idx <- complete.cases(best_exposure, benthic_sf$Shannon_Index)

# Fit model
best_model <- lm(Shannon_Index ~ best_exposure,
                 data = data.frame(
                   Shannon_Index = benthic_sf$Shannon_Index[valid_idx],
                   best_exposure = best_exposure[valid_idx]
                 ))

# Model summary
cat("\n========================================\n")
cat("BEST MODEL SUMMARY\n")
cat("Coefficient:", best_coef, "\n")
cat("========================================\n")
print(summary(best_model))

# Diagnostic plots
par(mfrow = c(2, 2))
plot(best_model, main = paste("Coefficient:", best_coef))
par(mfrow = c(1, 1))
```

# 10. Spatial Maps Comparison
```{r spatial_maps, fig.height=12}
par(mfrow = c(length(coefficients), 1), mar = c(4, 4, 3, 6))

for (i in seq_along(coefficients)) {
  coef <- coefficients[i]
  
  if (!is.null(cumulative_rasters[[as.character(coef)]])) {
    
    # Plot exposure map
    plot(cumulative_rasters[[as.character(coef)]],
         main = paste("Cumulative Exposure - Coefficient:", coef,
                     "| Correlation:", round(results$Correlation_Spearman[i], 3)),
         col = terrain.colors(100))
    
    # Overlay stations colored by Shannon Index
    points(st_coordinates(benthic_sf),
           pch = 21,
           bg = terrain.colors(100)[cut(benthic_sf$Shannon_Index, breaks = 100)],
           cex = 1.5,
           lwd = 1)
    
    # Highlight best coefficient
    if (coef == best_coef) {
      mtext("*** BEST MODEL ***", side = 3, line = 0.5, col = "red", font = 2, cex = 1.2)
    }
  }
}

par(mfrow = c(1, 1))
```

# 11. Save Best Cumulative Exposure
```{r save_best}
# Save the best cumulative exposure raster
best_cumulative <- cumulative_rasters[[as.character(best_coef)]]

if (!is.null(best_cumulative)) {
  writeRaster(best_cumulative, 
              "cumulative_exposure_BEST.tif",
              overwrite = TRUE)
  
  cat("✓ Saved best cumulative exposure to: cumulative_exposure_BEST.tif\n")
}

# Save benthic data with all exposure values
benthic_export <- benthic_sf %>%
  st_drop_geometry()

write.csv(benthic_export, 
          "benthic_with_exposures.csv",
          row.names = FALSE)

cat("✓ Saved benthic data with exposures to: benthic_with_exposures.csv\n")

# Save results table
write.csv(results, 
          "calibration_results.csv",
          row.names = FALSE)

cat("✓ Saved calibration results to: calibration_results.csv\n")
```

# 12. Conclusions
```{r conclusions}
cat("\n========================================\n")
cat("CALIBRATION CONCLUSIONS\n")
cat("========================================\n\n")

cat("Based on", nrow(benthic_sf), "sampling stations:\n\n")

cat("BEST COEFFICIENT:", best_coef, "\n")
cat("  - Spearman correlation:", round(best_cor, 4), "\n")
cat("  - R²:", round(results$R_squared[1], 4), "\n")
cat("  - P-value:", format.pval(results$P_value[1]), "\n\n")

if (abs(best_cor) > 0.5) {
  cat("✓ STRONG correlation between exposure and Shannon Index\n")
} else if (abs(best_cor) > 0.3) {
  cat("✓ MODERATE correlation between exposure and Shannon Index\n")
} else {
  cat("⚠ WEAK correlation between exposure and Shannon Index\n")
  cat("  Consider: - Adding more stressors\n")
  cat("           - Checking stressor locations\n")
  cat("           - Reviewing decay types\n")
}

cat("\nNext steps:\n")
cat("1. Use coefficient", best_coef, "in your final exposure model\n")
cat("2. Review spatial patterns in the maps above\n")
cat("3. Consider if results make ecological sense\n")
```



```{r test why}
# =================================================================
# CORRECTED DIAGNOSTICS
# =================================================================

library(ggplot2)

# Get data ready - FIX THE COLUMN NAME REFERENCE
benthic_df <- benthic_sf %>%
  st_drop_geometry() %>%
  mutate(
    Latitude = st_coordinates(benthic_sf)[,2],
    Longitude = st_coordinates(benthic_sf)[,1],
    Exposure = `Exposure_2.5e-05`  # Use backticks!
  )

# Alternative if backticks don't work:
# benthic_df$Exposure <- benthic_df[[paste0("Exposure_", 2.5e-05)]]

# =================================================================
# 1. WHERE ARE HIGH/LOW SHANNON VALUES?
# =================================================================
cat("\n========================================\n")
cat("SHANNON INDEX DISTRIBUTION\n")
cat("========================================\n")
summary(benthic_df$Shannon_Index)

# Which stations have lowest diversity?
low_diversity <- benthic_df %>%
  arrange(Shannon_Index) %>%
  head(10) %>%
  select(Station, Shannon_Index, Exposure, Sector, Depth)

cat("\nLowest Diversity Stations:\n")
print(low_diversity)

# Which stations have highest diversity?
high_diversity <- benthic_df %>%
  arrange(desc(Shannon_Index)) %>%
  head(10) %>%
  select(Station, Shannon_Index, Exposure, Sector, Depth)

cat("\nHighest Diversity Stations:\n")
print(high_diversity)

# =================================================================
# 2. EXPOSURE DISTRIBUTION
# =================================================================
cat("\n========================================\n")
cat("EXPOSURE DISTRIBUTION\n")
cat("========================================\n")
summary(benthic_df$Exposure)

cat("\nStations with HIGHEST exposure:\n")
high_exposure <- benthic_df %>%
  arrange(desc(Exposure)) %>%
  head(10) %>%
  select(Station, Shannon_Index, Exposure, Sector)
print(high_exposure)

cat("\nStations with LOWEST exposure:\n")
low_exposure <- benthic_df %>%
  arrange(Exposure) %>%
  head(10) %>%
  select(Station, Shannon_Index, Exposure, Sector)
print(low_exposure)

# =================================================================
# 3. DO HIGH EXPOSURE STATIONS HAVE LOW DIVERSITY?
# =================================================================
cat("\n========================================\n")
cat("CROSS-CHECK: Exposure vs Shannon\n")
cat("========================================\n")

# Split into exposure groups
benthic_df$Exposure_Group <- cut(benthic_df$Exposure, 
                                  breaks = 3, 
                                  labels = c("Low", "Medium", "High"))

# Compare Shannon by exposure group
cat("\nShannon Index by Exposure Level:\n")
shannon_by_exp <- benthic_df %>%
  group_by(Exposure_Group) %>%
  summarise(
    N = n(),
    Mean_Shannon = mean(Shannon_Index, na.rm = TRUE),
    SD_Shannon = sd(Shannon_Index, na.rm = TRUE),
    .groups = "drop"
  )
print(shannon_by_exp)

# =================================================================
# 4. WHAT ABOUT NATURAL FACTORS?
# =================================================================
cat("\n========================================\n")
cat("CORRELATIONS WITH ALL VARIABLES\n")
cat("========================================\n")

# Check all possible correlations
numeric_vars <- names(benthic_df)[sapply(benthic_df, is.numeric)]
correlations <- data.frame(
  Variable = character(),
  Correlation = numeric(),
  stringsAsFactors = FALSE
)

for (var in numeric_vars) {
  if (var != "Shannon_Index" && sum(!is.na(benthic_df[[var]])) > 10) {
    corr <- cor(benthic_df[[var]], benthic_df$Shannon_Index, 
                use = "complete.obs", method = "spearman")
    correlations <- rbind(correlations, 
                         data.frame(Variable = var, Correlation = corr,
                                   stringsAsFactors = FALSE))
  }
}

correlations <- correlations %>%
  arrange(desc(abs(Correlation)))

print(correlations)

cat("\n*** Variables with |correlation| > 0.3 are potentially important! ***\n")

# =================================================================
# 5. VISUALIZATIONS
# =================================================================

# Shannon vs Latitude
p1 <- ggplot(benthic_df, aes(x = Latitude, y = Shannon_Index)) +
  geom_point(aes(color = Exposure), size = 3) +
  geom_smooth(method = "loess", color = "red") +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Does Shannon Index change along the fjord?",
       subtitle = "Points colored by exposure level") +
  theme_minimal()

print(p1)

# Shannon vs Depth
if ("Depth" %in% names(benthic_df)) {
  p2 <- ggplot(benthic_df, aes(x = Depth, y = Shannon_Index)) +
    geom_point(aes(color = Exposure), size = 3) +
    geom_smooth(method = "loess", color = "blue") +
    scale_color_gradient(low = "blue", high = "red") +
    labs(title = "Is depth more important than exposure?") +
    theme_minimal()
  
  print(p2)
}

# Shannon by Sector
if ("Sector" %in% names(benthic_df)) {
  p3 <- ggplot(benthic_df, aes(x = Sector, y = Shannon_Index, fill = Sector)) +
    geom_boxplot() +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3) +
    labs(title = "Shannon Index by Sector",
         subtitle = "Diamond = mean") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p3)
}

# Exposure vs Shannon scatter with sector colors
if ("Sector" %in% names(benthic_df)) {
  p4 <- ggplot(benthic_df, aes(x = Exposure, y = Shannon_Index, color = Sector)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    labs(title = "Shannon Index vs Cumulative Exposure",
         subtitle = "Colored by sector - showing weak relationship") +
    theme_minimal()
  
  print(p4)
}

# =================================================================
# 6. CHECK INDIVIDUAL STRESSORS
# =================================================================

cat("\n========================================\n")
cat("INDIVIDUAL STRESSOR ANALYSIS\n")
cat("========================================\n")

individual_results <- data.frame(
  Stressor = character(),
  Correlation = numeric(),
  Mean_Exposure = numeric(),
  stringsAsFactors = FALSE
)

for (stressor in c("agriculture", "coastal_development", "ice_fishing", "wwtp", "forestry")) {
  
  exp_rast <- load_exposure_raster(stressor, 0.000025)
  
  if (!is.null(exp_rast)) {
    # Normalize
    max_val <- cellStats(exp_rast, "max", na.rm = TRUE)
    if (!is.na(max_val) && max_val > 0) {
      exp_rast <- exp_rast / max_val
    }
    
    # Extract
    exp_vals <- raster::extract(exp_rast, st_coordinates(benthic_sf))
    
    # Stats
    valid <- complete.cases(exp_vals, benthic_sf$Shannon_Index)
    if (sum(valid) > 10) {
      corr <- cor(exp_vals[valid], benthic_sf$Shannon_Index[valid], method = "spearman")
      mean_exp <- mean(exp_vals[valid], na.rm = TRUE)
      
      individual_results <- rbind(individual_results,
                                 data.frame(Stressor = stressor,
                                          Correlation = corr,
                                          Mean_Exposure = mean_exp,
                                          stringsAsFactors = FALSE))
    }
  }
}

individual_results <- individual_results %>%
  arrange(desc(abs(Correlation)))

cat("\nIndividual Stressor Correlations:\n")
print(individual_results)

# Plot
if (nrow(individual_results) > 0) {
  barplot(individual_results$Correlation,
          names.arg = individual_results$Stressor,
          las = 2,
          col = ifelse(individual_results$Correlation < 0, "red", "blue"),
          main = "Individual Stressor Correlations",
          ylab = "Spearman Correlation",
          ylim = c(min(individual_results$Correlation) - 0.1,
                   max(individual_results$Correlation) + 0.1))
  abline(h = 0, lty = 2)
  abline(h = c(-0.3, 0.3), lty = 2, col = "green")
  text(x = 1, y = 0.3, "Moderate correlation", pos = 3, col = "green", cex = 0.8)
}
```
```{r try calibration with abumdance }
# TEST WITH TOTAL ABUNDANCE INSTEAD OF SHANNON INDEX
library(raster)
library(sf)
library(dplyr)
library(ggplot2)

# Reload benthic data
benthic_df <- benthic_sf %>%
  st_drop_geometry() %>%
  mutate(
    Latitude = st_coordinates(benthic_sf)[,2],
    Longitude = st_coordinates(benthic_sf)[,1]
  )

cat("========================================\n")
cat("CALIBRATION WITH TOTAL ABUNDANCE\n")
cat("========================================\n\n")

# Check abundance distribution
cat("Total Abundance Distribution:\n")
summary(benthic_df$Total_Abundance)
hist(benthic_df$Total_Abundance, breaks = 20, 
     main = "Distribution of Total Abundance",
     xlab = "Total Abundance", col = "lightblue")

# Test all three existing coefficients
abundance_results <- data.frame(
  Coefficient = numeric(),
  Correlation_Spearman = numeric(),
  Correlation_Pearson = numeric(),
  R_squared = numeric(),
  P_value = numeric()
)

for (coef in c(0.00001, 0.00005, 0.000025)) {
  
  cat("\n--- Coefficient:", coef, "---\n")
  
  # Get exposure for this coefficient
  exposure_col <- paste0("Exposure_", format(coef, scientific = TRUE))
  
  if (exposure_col %in% names(benthic_df)) {
    
    exposure <- benthic_df[[exposure_col]]
    abundance <- benthic_df$Total_Abundance
    
    # Remove NAs
    valid_idx <- complete.cases(exposure, abundance)
    
    if (sum(valid_idx) > 10) {
      
      # Correlations
      spearman <- cor(exposure[valid_idx], abundance[valid_idx], 
                      method = "spearman")
      pearson <- cor(exposure[valid_idx], abundance[valid_idx], 
                     method = "pearson")
      
      # Linear model
      model <- lm(abundance ~ exposure, 
                  data = data.frame(abundance = abundance[valid_idx],
                                   exposure = exposure[valid_idx]))
      
      r2 <- summary(model)$r.squared
      pval <- summary(model)$coefficients[2, 4]
      
      abundance_results <- rbind(abundance_results, data.frame(
        Coefficient = coef,
        Correlation_Spearman = spearman,
        Correlation_Pearson = pearson,
        R_squared = r2,
        P_value = pval
      ))
      
      cat("Spearman correlation:", round(spearman, 4), "\n")
      cat("R²:", round(r2, 4), "\n")
      cat("P-value:", format.pval(pval), "\n")
    }
  }
}

# Results table
cat("\n========================================\n")
cat("RESULTS SUMMARY - ABUNDANCE\n")
cat("========================================\n")
print(abundance_results)

# Find best
best_idx <- which.max(abs(abundance_results$Correlation_Spearman))
cat("\nBest coefficient for ABUNDANCE:", 
    abundance_results$Coefficient[best_idx], "\n")
cat("Correlation:", 
    round(abundance_results$Correlation_Spearman[best_idx], 4), "\n")

# ================================================================
# VISUALIZE: ABUNDANCE VS EXPOSURE
# ================================================================

# Use best coefficient
best_coef <- abundance_results$Coefficient[best_idx]
best_exposure_col <- paste0("Exposure_", format(best_coef, scientific = TRUE))

benthic_df$Best_Exposure <- benthic_df[[best_exposure_col]]

# Scatter plot
ggplot(benthic_df, aes(x = Best_Exposure, y = Total_Abundance)) +
  geom_point(aes(color = Sector), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  labs(
    title = "Total Abundance vs Cumulative Exposure",
    subtitle = paste("Coefficient:", best_coef, 
                     "| Correlation:", 
                     round(abundance_results$Correlation_Spearman[best_idx], 3)),
    x = "Cumulative Exposure",
    y = "Total Abundance (individuals)"
  ) +
  theme_minimal(base_size = 14)

# ================================================================
# COMPARE: SHANNON VS ABUNDANCE
# ================================================================

cat("\n========================================\n")
cat("COMPARISON: Shannon Index vs Abundance\n")
cat("========================================\n\n")

# For the same coefficient, which response variable is better?
comparison <- data.frame(
  Coefficient = c(0.00001, 0.00005, 0.000025),
  Shannon_Correlation = c(-0.093, -0.027, -0.102),  # From earlier
  Abundance_Correlation = abundance_results$Correlation_Spearman
)

print(comparison)

# Which is more sensitive?
cat("\n")
if (max(abs(comparison$Abundance_Correlation)) > 
    max(abs(comparison$Shannon_Correlation))) {
  cat("✓
```
```{r comparison}
# ================================================================
# COMPLETE COMPARISON
# ================================================================

cat("\n========================================\n")
cat("COMPARISON: Shannon Index vs Abundance\n")
cat("========================================\n\n")

# Comparison table
comparison <- data.frame(
  Coefficient = c(0.00001, 0.00005, 0.000025),
  Shannon_Correlation = c(-0.093, -0.027, -0.102),
  Abundance_Correlation = c(0.1283, 0.1246, 0.1164)
)

print(comparison)

# Which is more sensitive?
cat("\n")
if (max(abs(comparison$Abundance_Correlation)) > 
    max(abs(comparison$Shannon_Correlation))) {
  cat("✓ ABUNDANCE shows STRONGER correlation than Shannon Index!\n\n")
} else {
  cat("✗ Shannon Index shows stronger correlation than Abundance\n\n")
}

# The key insight
cat("========================================\n")
cat("KEY FINDING:\n")
cat("========================================\n")
cat("Exposure shows OPPOSITE effects:\n")
cat("  • Shannon Index: NEGATIVE correlation (-0.10)\n")
cat("  • Abundance:     POSITIVE correlation (+0.13)\n\n")
cat("Interpretation:\n")
cat("High exposure areas have:\n")
cat("  ✓ MORE individuals (organic enrichment)\n")
cat("  ✓ FEWER species (pollution-intolerant species lost)\n\n")
cat("This is a CLASSIC organic pollution signature!\n")
cat("========================================\n")

# ================================================================
# VISUALIZE BOTH METRICS TOGETHER
# ================================================================

library(gridExtra)

# Shannon vs Exposure
p1 <- ggplot(benthic_df, aes(x = `Exposure_1e-05`, y = Shannon_Index)) +
  geom_point(aes(color = Sector), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Shannon Index vs Exposure",
    subtitle = "Negative trend (r = -0.09)",
    x = "Cumulative Exposure",
    y = "Shannon Diversity Index"
  ) +
  theme_minimal(base_size = 12)

# Abundance vs Exposure
p2 <- ggplot(benthic_df, aes(x = `Exposure_1e-05`, y = Total_Abundance)) +
  geom_point(aes(color = Sector), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(
    title = "Total Abundance vs Exposure",
    subtitle = "Positive trend (r = +0.13)",
    x = "Cumulative Exposure",
    y = "Total Abundance"
  ) +
  theme_minimal(base_size = 12)

# Side by side
grid.arrange(p1, p2, ncol = 2,
             top = "OPPOSITE EFFECTS: Organic Enrichment Pattern")

# ================================================================
# CHECK HIGH EXPOSURE STATIONS IN DETAIL
# ================================================================

cat("\n========================================\n")
cat("HIGH EXPOSURE STATIONS - DETAILED VIEW\n")
cat("========================================\n\n")

high_exp <- benthic_df %>%
  arrange(desc(`Exposure_1e-05`)) %>%
  head(10) %>%
  select(Station, Shannon_Index, Total_Abundance, 
         Species_Richness, Exposure = `Exposure_1e-05`, Sector)

print(high_exp)

cat("\nPattern Check:\n")
cat("Mean Shannon (high exposure):", 
    mean(high_exp$Shannon_Index), "\n")
cat("Mean Abundance (high exposure):", 
    mean(high_exp$Total_Abundance), "\n")

low_exp <- benthic_df %>%
  filter(`Exposure_1e-05` < 0.001) %>%
  summarise(
    Mean_Shannon = mean(Shannon_Index),
    Mean_Abundance = mean(Total_Abundance)
  )

cat("\nMean Shannon (low exposure):", 
    low_exp$Mean_Shannon, "\n")
cat("Mean Abundance (low exposure):", 
    low_exp$Mean_Abundance, "\n")
```
